We are not the 99 percent: quantifying asphericity in the distribution of Local Group satellites

### 2018-01-24 paper/paper.tex
edit

\begin{abstract}
We quantify the asphericity in the spatial distribution of the
brightest satellites around the Milky Way (MW) and M31, the dominant
galaxies in the Local Group (LG).
We use simulations to build an explicit probability distribution for the
asphericity around galaxies similar to the LG in the Lambda Cold Dark
Matter (LCDM) paradigm.
This allows us to estimate the atipicallity
of the satellite distributions in the LG even when the underlying
simulations do not have enough systems that fully resemble the LG.
We demonstrate the method using three different simulations:
Illustris-1,  Illustris-1-DarkMatter and ELVIS. 
We find three common trends in these simulations. First, at most of
$1\%$ the pairs are expected to have satellite distributions with the
same degree of deviation from sphericity as the Local Group; second, between
$38\%$ to $59\%$ of the pairs have a halo with a satellite
distribution as aspherical as  in M31; and third, at most $2\%$ of the
pairs have a satellite distribution as planar as in the MW. 
The M31 galaxy turns out to have a rather typical distribution 
in the LCDM context, for its brightest satellites, while the MW is
atypical in this respect. 
These results place the LG close to the level of a 3-$\sigma$ outlier
in the LCDM paradigm.
We summarize our results in terms of mean values and covariance
matrices. This approach will facilitate the direct comparison of the
results from different numerical setups and choices to select LG pairs
in simulations. 
\end{abstract}

@@

The presence of a Vast Polar Structure of satellites around the Milky
Way has been stablished in the last decade.
This structure can be easily described by global quantities of the
brightest

The question whether this configuration is atypical in the observable
universe cannot be proberly addressed given the current limits in
observations. 
So far it has only been possible to contraint that $5\%$ of the
Milky-Way like galaxies in the Universe have two bright satellite
galaxies as is the case of the Milky Way and its Large and Small
Magellanic Clouds.

The closest system that could help us in providing a baseline for
comparison is the Andromeda galaxy, M31. 
Its close enough that an observational study of its satellite
population is possible, with the great advantage of having a
perpective from a distance and not from within, as is the case for our
Galaxy. 
These studies also found a planar satellite
distributions.
This time not in the whole distribution, but in a subset of $15$
satellites out of an homogeneous sample of $27$ satellites.

This has prompted a discussion on whether this satellite distributions
are easily found in structure formation simulations done in the
current dominant paradigm for galaxy formation.
This paradigm, based on a Cold Dark Matter cosmology in
a expanding universe described by General Relativity with a
Cosmological Constant, is the so-called Lambda Cold Dark Matter
(LCDM).

The consensus is that these structures are hard to find in
simulations. 
Some studies have focused on the study of individual dark matter halos
that could host galaxies like the MW or M31, while some other studies
have tried to simulate the formation of galaxy pairs resembling both
the MW and M31.
The transition from studying individual halos to pairs has been
motivated by the hypotesis hat the location of the Local Group in
the cosmic web should play a determinant role in building up planar
satellite distributions through preferential alignments and accretions
histories of galaxies.

>>>>

\begin{abstract}
We quantify the asphericity in the spatial distribution of the
brightest satellites around the Milky Way (MW) and M31, the dominant
galaxies in the Local Group (LG).
We use simulations to build an explicit probability distribution for the
asphericity around galaxies similar to the LG in the Lambda Cold Dark
Matter (LCDM) paradigm.
This allows us to estimate the atipicallity
of the satellite distributions in the LG even when the underlying
simulations do not have enough systems that fully resemble the LG.
We demonstrate the method using three different simulations:
Illustris-1,  Illustris-1-DarkMatter and ELVIS. 
Detailed results differ greatly among the simulations, suggesting a
strong influence of resolution, modelled physics and criteria to select
LG pairs.
However, there are three common trends in these simulations. First, at most of
$1\%$ the pairs are expected to have satellite distributions with the
same asphericity as the Local Group; second, between
$38\%$ to $59\%$ of the pairs have a halo with a satellite
distribution as aspherical as in M31; and third, at most $2\%$ of the
pairs have a satellite distribution as planar as in the MW. 
The brightest satellites in M31 galaxy turn out to have a rather typical distribution 
in the LCDM context, while the MW is atypical in this respect. 
These results place the LG close to the level of a 3-$\sigma$ outlier
in the LCDM paradigm.
We summarize our results in terms of mean values and covariance
matrices. 
This approach facilitates the comparison between asphericity results
from different numerical setups and choices to select LG pairs in
simulations.  
\end{abstract}

@@

The spatial distribution of satellite galaxies around our Milky Way and the
M31 galaxy provides a stringent test for structure formation theories
in a explicit cosmological context. 
The presence of a Vast Polar Structure of satellites around the Milky
Way has been stablished in the last decade.
This structure is present in the $11$ brightest "classical" dwarf
galaxies and has been confirmed as fainter satellites have been
detected. 
In the case of M31,  observational studies have found a planar satellite
distribution in a subset of $15$ satellites out of an homogeneous
sample of $27$ satellites. 
Using only the $11$ brightest satellites (as a comparison with the MW)
there does not seem to be any structure.

The question whether this configuration is atypical in the observable
Universe cannot be proberly addressed given the current limits in
observations. 
So far it has only been possible to contraint that $5\%$ of the
Milky-Way like galaxies in the Universe have two bright satellite
galaxies as is the case of the Milky Way and its Large and Small
Magellanic Clouds.

However, the tipicallity of satellite distributions can be addressed
in simulations.
Most of the studies have been performed in the currently preferred
paradigm, which is based on a Cold Dark Matter cosmology in a expanding
Universe described by General Relativity including a Cosmological
Constant, the so-called Lambda Cold Dark Matter (LCDM).
The broad consensus is that these highly structured satellite distributions
are hard to find in simulations.  
There are still disagreements on the precise degree of atypicallity
and the conditions required to simulate

Some studies have focused on the study of individual dark matter halos
that could host galaxies like the MW or M31, while some other studies
have tried to simulate the formation of galaxy pairs resembling both
the MW and M31.
The transition from studying individual halos to pairs has been
motivated by the hypotesis hat the location of the Local Group in
the cosmic web should play a determinant role in building up planar
satellite distributions through preferential alignments and accretions
histories of galaxies.
